Mechanisms of detachment in fibrillar adhesive systems

### 2021-06-03 source/tex/manuscript.tex
Final edits: "fibrillar", new references added

\title{Mechanisms of detachment in hairy adhesive systems}

@@

\begin{abstract}
Several creatures can climb on smooth surfaces with the help of hairy
adhesive pads on their legs. A rapid change from strong attachment
to effortless detachment of the leg is enabled by the asymmetric geometry
of the hairs. In this study, we propose possible mechanisms by which
a hairy pad can be easily detached even when the hairs possess no
asymmetry. Based on a spring-based model, we consider three modes
of detachment: vertically pulling the pad while maintaining either
a 1) fixed or a 2) free leg joint, or by 3) flexing the pad about
the claw. We show that in all cases, the adhesion force can be significantly
reduced due to elastic forces when the hairs deform non-uniformly
across the array, an effect that we call ``\emph{elastic weakening}''.
Our proposed model illustrates the design advantage provided by hairy
adhesive systems, which not only provides strong adhesion, but also
allows easy detachment, making it suitable as organs for fast locomotion.
The presented approaches can be potentially used to switch the adhesion
state in bio-inspired artificial fibrillar adhesives without the need of asymmetric
structures.
\end{abstract}

@@

Over the past few decades, there have been numerous studies to understand
how animals such as geckos and insects are able to walk on surfaces
of any direction while defying gravity. A microscopic observation
reveals that, in many cases, animals have a dense array of hairy structures
at the end of their legs \cite{RN198,RN59}. These hairy adhesive
pads help the animal to stay attached to any surface or detach easily
at will for countless cycles, a property that is referred to as \emph{reversible
adhesion}. Previous attempts to theoretically explain adhesion in
hairy pads \cite{RN40,RN120} has followed two fundamental approaches:
either by energy balance, or by force balance.

In the energy balance approach, adhesion is usually characterized by ``work
of adhesion'' (W\textsubscript{adh}), which is the energy required
to separate a pad from the surface. During detachment, the elastic
energy stored in the hair is dissipated, which increases W\textsubscript{adh}
and thus adhesion is enhanced \cite{RN29,RN114}. Detachment of an
individual hair can be explained based on Kendall's peeling theory \cite{RN147},
which predicts low forces at high peeling angles.

@@

The theories presented so far suggests that low detachment force of
a hairy pad can be attained by 1) increasing the stress concentration
by peeling the pad at high angles, or 2) laterally shearing the pad
before pull-off, which requires the hairs to have an asymmetric geometry
or curvature. However, some insects like male dock beetles predominantly
possess flat ``mushroom'' tipped hairs on their pads \cite{RN19},
which are relatively less asymmetric compared to the ``spatula''
tipped hairs. This begs the question of how the beetle can still easily
detach their legs during locomotion, since these mushroom shaped hairs
should be resilient to detachment via lateral shearing. Artificial
micro-pillar adhesives mimicking the mushroom shaped hairs of the
beetle have in fact demonstrated its superior adhesive strength in
comparison to other geometries \cite{RN112,RN57}. Introducing asymmetry
into the pillar geometry of such mimics for easy detachment is however
challenging due to current limitations in fabrication techniques and
difficulty in scaling-up for practical applications \cite{RN57}.
Alternate strategies are thus desired to switch the adhesion state
of symmetric pillar arrays in a reversible manner, e.g. by buckling
the pillars under compressive load leading to their contact loss \cite{RN279}
or by using special materials reacting to external stimuli such as
magnetic field \cite{RN186}, UV light \cite{RN280} or temperature \cite{RN281}.

In this work, we propose the possible mechanisms by which adhesive
pads with symmetric hairs can be easily detached under a theoretical
framework. The hairy pad is modelled as an array of springs and the
force-balance approach is used to quantify the net adhesion \cite{RN117}.
We hope our work to provide strategies to control the adhesion force
of an artificial micro-pillar adhesive, which has applications in
bio-inspired climbing robots, pick-and-place operations and reusable
adhesives.

@@

The hairy pad is assumed to be a one dimensional array of $N_{t}$
hairs, each behaving like a spring with spring constant, $k_{h}$,
and natural length, $l_{h,0}$ (Figure \ref{fig:Schematic}). The
array backing is assumed to be stiff. The pad is attached to a linearly
deformable leg, assumed to be another spring with spring constant,
$k_{l}$, and natural length, $l_{l,0}$. The leg is hinged to the
array at a distance, $s$, from the right end of the array. The hinge
is at a vertical distance, $d_{s}$, from the surface. The hairs are
spaced apart by a width, $w$, and the array is of length, $L=\left(N_{t}-1\right)w$.
The pad is oriented at an instantaneous angle, $\theta$, while making contact with
a flat smooth surface. Each hair can attain a maximum length, $l_{h,p}$,
before pull-off, such that its pull-off force, $f_{p}=k_{h}\left(l_{h,p}-l_{h,0}\right)$.
$F_{net}$ is the net force on the pad and $M_{net}$ is the the net
torque about the hinge, at a particular instant during the detachment
process.

\begin{figure}[H]
\includegraphics{Figure1-model_schematic}\caption{Spring contact model of a hairy adhesive pad. The pad consists of
an array of $N_{t}$ hairs connected to a deformable leg at the hinge.
At a particular distance, $d_{s}$, $n$ number of hairs are in contact
and the array is oriented at a tilt angle, $\theta$, with the surface.
\label{fig:Schematic}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics{Case1-Fixed_pull}\caption{Force distance curves for a hairy pad, pulled vertically upwards with
a fixed hinge between the pad and the leg (\emph{Fixed pull}). The
tilt angle, $\theta$, of the array is kept fixed during detachment.
The leg is either stiff ($k_{l}\rightarrow\infty$) or soft ($k_{l}/k_{h}=10$).
Positive force values represent attraction. The green dashed line
represents the maximum possible adhesion for the pad. All values are
normalized to dimensionless forms, as described in text.\label{fig:Fixed-pull}}
\end{figure}

The hairy adhesive pad can be detached by pulling it vertically upwards
while maintaining a constant tilt angle, $\theta$, with the surface.
This can be achieved if the joint connecting the leg to the array
(leg-hinge) is kept fixed. Equations \ref{eq:force_n_d}, \ref{eq:d_max}
and \ref{eq:simul_detach} can be used to get the resulting force-distance
curves.

@@

For the case of a soft leg, we observe a similar effect of tilt angle
on the force curves as before. The maximum adhesion force at a particular
tilt is the same as that for the stiff leg. The saw-tooth jitter are
however minimized due to the leg's deformation, leading to a ``dampened''
force response. Interestingly, the force abruptly drops to zero for
the angles $0.2\lyxmathsym{\textdegree}$ and $0.4\lyxmathsym{\textdegree}$.
This is an effect of the elastic recoil of the leg while each hair
loses contact during the detachment process (equation \ref{eq:backing_deform}).
The length difference between the detached hair just before it breaks
contact and its adjacent hair is $w\sin\theta$. If the leg's recoil
length, $\Delta l>w\sin\theta$, the adjacent hair will be stretched
more than its maximum length ($l_{h,p}$), and thus will also detach,
leading to further recoil of the leg. Equation \ref{eq:backing_deform}
shows that $\Delta l$ increases with every subsequent loss of hair
contact if $\theta$ is kept constant. This implies that, once initiated,
the leg's recoil will always be large enough to detach every remaining
hair, resulting in a spontaneous propagation of the detachment front
until the pad completely breaks contact with the surface.

@@

\includegraphics{Case2-Free_pull}\caption{Force distance curves for a hairy pad, pulled vertically upwards with
a free hinge between the pad and the leg (\emph{Free pull}). $s$
is the distance between the leg-hinge and the right end of the array
and $L$ is the array length. The free hinge allows further tilting
of the array during the vertical pull. The leg is either stiff ($k_{l}\rightarrow\infty$)
or soft ($k_{l}/k_{h}=10$). Positive force values represent attraction.
The green dashed line represents the maximum possible adhesion for
the pad. All values are normalized to dimensionless forms, as described
in text. \label{fig:Free-pull}}

@@

\begin{figure}[H]
\includegraphics{Case3-Fixed_flex}\caption{Force curves for a hairy pad detached by flexing it about the claw.
$\hat{F}_{pull}=\frac{F_{pull}}{k_{h}w}$ is the normalised pulling
force necessary to apply the torque about the claw-hinge for detachment,
$\hat{F}_{hinge}=\frac{F_{hinge}}{k_{h}w}$ is the normalised reaction
force on the claw-hinge, $\hat{d_{s}}=\frac{d_{s}-l_{h,0}}{w}$ is
the normalized vertical distance of the claw-hinge from the surface.
Here, $s_{l}/L=1$ and $s_{h}/w=10$. The green dashed line represents
the maximum possible adhesion for the pad. \label{fig:Flex}}
\end{figure}

@@

Based on the three modes of detachment discussed in the previous sections,
one can think of several strategies to detach a hairy adhesive pad
from the surface. To illustrate this, let us consider the adhesive
system of a dock beetle. The beetle is known to have 3 sets of hairy
pads in each of their legs, each possessing hairs of different geometries.
To keep our analysis simple, we will assume each leg to have only
two pads, with identical hairs of ``mushroom'' geometry. The distal
and proximal pads possess roughly 500 and 1000 hairs, respectively \cite{RN19}.
Assuming the pads to be rectangular arrays of $20\times25$ and $40\times25$
hairs, we can model this as a one-dimensional system of 20 and 40
hairs, respectively, by combining the hairs along the width. Based
on reported measurements \cite{RN79}, the beetle's ``combined hair''
is thus considered to have an effective pull-off force, $f_{p}=0.5\times25=12.5$
\textgreek{m}N and spring constant, $k_{h}=0.5\times25=12.5$ N m\protect\textsuperscript{-1}.
The beetle's hairs are approximately $l_{h,0}=40$ \textgreek{m}m
long, spaced $w=10$ \textgreek{m}m apart. The adhesive pad also has
a claw, around 200 \textgreek{m}m long, and the leg is connected roughly
at the end of the proximal pad. This will put $\hat{s}_{h}=20$ and
$s_{l}/L=1$, measured relative to the right end of the distal pad.
The beetle's leg is assumed to possess two joints which could serve
as a hinges for rotation (H\textsubscript{1} and H\textsubscript{2}
in Figure \ref{fig:Detachment-pathways} inset). The claw can be used
as an external hinge (H\textsubscript{3}) by flexing the pad around
it.

@@

Inhomogeneous deformation of hairs results in significant reductions
in the adhesion of a hairy pad system. Such a condition can be achieved
by either 1) pulling the pad while maintaining a constant tilt angle,
2) pulling the pad while maintaining a free leg joint or 3) flexing
the pad around the claw. Strategic control of the joint's mobility
can allow the leg to easily switch between the above mechanisms, thus
providing a simple way to reduce adhesion as per necessity. The presence
of a deformable leg can further trigger a spontaneous propagation
of hair detachment due to the leg's elastic recoil, making it suitable
for fast detachment. Arrays with low $\hat{f}_{p}$ and large number
of hairs, with a hair geometry that allows for large deformations
while avoiding buckling and lateral bundling represent the optimal
design conditions to maximize the range of control over adhesion.
The proposed theory is supported by previously reported experimental
observations of leg detachment in dock beetles and opens up new strategies
to achieve reversible adhesion in bio-inspired artificial fibrillar adhesives
without the need of asymmetric structures.

@@

We are grateful to Hans-J\"{u}rgen Butt and Thomas Endlein for reviewing the text, and thank Ren\textcyr{\`\cyre} Hensel and
Bat-El Pinchasik for fruitful discussions. This work was supported by the Deutsche Forschungsgemeinschaft
[Grant number: PI 1351/2-1] and the Max Planck Graduate Center with the Johannes Gutenberg-Universit\"{a}t Mainz [MPGC].

>>>>

\title{Mechanisms of detachment in fibrillar adhesive systems}

@@

\begin{abstract}
Several creatures can climb on smooth surfaces with the help of hairy
adhesive pads on their legs. A rapid change from strong attachment
to effortless detachment of the leg is enabled by the asymmetric geometry
of the hairs. In this study, we propose mechanisms by which
a hairy pad can be easily detached, even when the hairs possess no
asymmetry. Here, we examine the possible function of the leg joint and the claws.
Based on a spring-based model, we consider three modes
of detachment: vertically pulling the pad while maintaining either
a 1) fixed or a 2) free leg joint, or by 3) flexing the pad about
the claw. We show that in all cases, the adhesion force can be significantly
reduced due to elastic forces when the hairs deform non-uniformly
across the array, an effect that we call ``\emph{elastic weakening}''.
Our proposed model illustrates the design advantage provided by such fibrillar
adhesive systems, which not only provides strong adhesion, but also
allows easy detachment, making it suitable as organs for fast locomotion.
The presented approaches can be potentially used to switch the adhesion
state in bio-inspired fibrillar adhesives, by incorporating artificial joints and claws into their design, 
without the need of asymmetric or stimuli-responsive fibrillar structures.
\end{abstract}

@@

Over the past few decades, there have been numerous studies to understand
how animals such as geckos and insects are able to walk on surfaces
of any direction while defying gravity. A microscopic observation
reveals that, in many cases, animals have a dense array of fibrillar structures
at the end of their legs \cite{RN198,RN59}. These "hairy" adhesive
pads help the animal to stay attached to any surface or detach easily
at will for countless cycles, a property that is referred to as \emph{reversible
adhesion}. Previous attempts to theoretically explain adhesion in
hairy pads \cite{RN40,RN120} has followed two fundamental approaches:
either by energy balance, or by force balance.

In the energy balance approach, adhesion is usually characterized by ``work
of adhesion'' (W\textsubscript{adh}), which is the energy required
to separate a pad from the surface. During detachment, the elastic
energy stored in the hair is dissipated, which increases W\textsubscript{adh}
and thus adhesion is enhanced \cite{RN29,RN114}. Detachment of an
individual hair can be explained based on Kendall's peeling theory \cite{RN147, RN296},
which predicts low forces at high peeling angles.

@@

The theories presented so far suggests that low detachment force of
a fibrillar adhesive pad can be attained by 1) increasing the stress concentration
by peeling the pad at high angles, or 2) laterally shearing the pad
before pull-off, which requires the hairs to have an asymmetric geometry
or curvature. However, some insects like male dock beetles predominantly
possess flat ``mushroom'' tipped hairs on their pads \cite{RN19},
which are relatively less asymmetric compared to the ``spatula''
tipped hairs. This begs the question of how the beetle can still easily
detach their legs during locomotion, since these mushroom shaped hairs
should be resilient to detachment via lateral shearing. Artificial
micro-pillar adhesives mimicking the mushroom shaped hairs of the
beetle have in fact demonstrated its superior adhesive strength in
comparison to other geometries \cite{RN112,RN57}. Introducing asymmetry
into the pillar geometry of such mimics for easy detachment is however
challenging due to current limitations in fabrication techniques and
difficulty in scaling-up for practical applications \cite{RN57}.
Alternate strategies are thus desired to switch the adhesion state
of symmetric pillar arrays in a reversible manner, e.g. by buckling
the pillars under compressive load leading to their contact loss \cite{RN279}
or by using special materials reacting to external stimuli such as
magnetic field \cite{RN186}, UV light \cite{RN280} or temperature \cite{RN281}.

In this work, we propose the possible mechanisms by which adhesive
pads with symmetric hairs can be easily detached under a theoretical
framework. The hairy pad is modelled as an array of springs and the
force-balance approach is used to quantify the net adhesion \cite{RN117, RN253}.
We hope our work to provide strategies to control the adhesion force
of an artificial micro-pillar adhesive, which has applications in
bio-inspired climbing robots, pick-and-place operations and reusable
adhesives.

@@

The fibrillar adhesive pad is assumed to be a one dimensional array of $N_{t}$
hairs, each behaving like a spring with spring constant, $k_{h}$,
and natural length, $l_{h,0}$ (Figure \ref{fig:Schematic}). The
array backing is assumed to be stiff. The pad is attached to a linearly
deformable leg, assumed to be another spring with spring constant,
$k_{l}$, and natural length, $l_{l,0}$. The leg is hinged to the
array at a distance, $s$, from the right end of the array. The hinge
is at a vertical distance, $d_{s}$, from the surface. The hairs are
spaced apart by a width, $w$, and the array is of length, $L=\left(N_{t}-1\right)w$.
The pad is oriented at an instantaneous angle, $\theta$, while making contact with
a flat smooth surface. Each hair can attain a maximum length, $l_{h,p}$,
before pull-off, such that its pull-off force, $f_{p}=k_{h}\left(l_{h,p}-l_{h,0}\right)$.
$F_{net}$ is the net force on the pad and $M_{net}$ is the the net
torque about the hinge, at a particular instant during the detachment
process.

\begin{figure}[H]
\includegraphics{Figure1-model_schematic}\caption{Spring contact model of a fibrillar adhesive pad. The pad consists of
an array of $N_{t}$ hairs connected to a deformable leg at the hinge.
At a particular distance, $d_{s}$, $n$ number of hairs are in contact
and the array is oriented at a tilt angle, $\theta$, with the surface.
\label{fig:Schematic}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics{Case1-Fixed_pull}\caption{Force distance curves for a fibrillar adhesive pad, pulled vertically upwards with
a fixed hinge between the pad and the leg (\emph{Fixed pull}). The
tilt angle, $\theta$, of the array is kept fixed during detachment.
The leg is either stiff ($k_{l}\rightarrow\infty$) or soft ($k_{l}/k_{h}=10$).
Positive force values represent attraction. The green dashed line
represents the maximum possible adhesion for the pad. All values are
normalized to dimensionless forms, as described in text.\label{fig:Fixed-pull}}
\end{figure}

The fibrillar adhesive pad can be detached by pulling it vertically upwards
while maintaining a constant tilt angle, $\theta$, with the surface.
This can be achieved if the joint connecting the leg to the array
(leg-hinge) is kept fixed. Equations \ref{eq:force_n_d}, \ref{eq:d_max}
and \ref{eq:simul_detach} can be used to get the resulting force-distance
curves.

@@

For the case of a soft leg, we observe a similar effect of tilt angle
on the force curves as before. The maximum adhesion force at a particular
tilt is the same as that for the stiff leg. The saw-tooth jitter are
however minimized due to the leg's deformation, leading to a ``dampened''
force response. Interestingly, the force abruptly drops to zero for
the angles $0.2\lyxmathsym{\textdegree}$ and $0.4\lyxmathsym{\textdegree}$.
This is an effect of the elastic recoil of the leg while each hair
loses contact during the detachment process (equation \ref{eq:backing_deform}).
The length difference between the detached hair just before it breaks
contact and its adjacent hair is $w\sin\theta$. If the leg's recoil
length, $\Delta l>w\sin\theta$, the adjacent hair will be stretched
more than its maximum length ($l_{h,p}$), and thus will also detach,
leading to further recoil of the leg. Equation \ref{eq:backing_deform}
shows that $\Delta l$ increases with every subsequent loss of hair
contact if $\theta$ is kept constant. This implies that, once initiated,
the leg's recoil will always be large enough to detach every remaining
hair, resulting in a spontaneous propagation of the detachment front
until the pad completely breaks contact with the surface. This is consistent with 
a recent report of catastrophic failure, due to a similar recoil effect of the measurement system,
seen in micro-fibrillar adhesives with a narrow variance of individual fibril adhesive strengths \cite{RN273}.

@@

\includegraphics{Case2-Free_pull}\caption{Force distance curves for a fibrillar adhesive pad, pulled vertically upwards with
a free hinge between the pad and the leg (\emph{Free pull}). $s$
is the distance between the leg-hinge and the right end of the array
and $L$ is the array length. The free hinge allows further tilting
of the array during the vertical pull. The leg is either stiff ($k_{l}\rightarrow\infty$)
or soft ($k_{l}/k_{h}=10$). Positive force values represent attraction.
The green dashed line represents the maximum possible adhesion for
the pad. All values are normalized to dimensionless forms, as described
in text. \label{fig:Free-pull}}

@@

\begin{figure}[H]
\includegraphics{Case3-Fixed_flex}\caption{Force curves for a fibrillar adhesive pad detached by flexing it about the claw.
$\hat{F}_{pull}=\frac{F_{pull}}{k_{h}w}$ is the normalised pulling
force necessary to apply the torque about the claw-hinge for detachment,
$\hat{F}_{hinge}=\frac{F_{hinge}}{k_{h}w}$ is the normalised reaction
force on the claw-hinge, $\hat{d_{s}}=\frac{d_{s}-l_{h,0}}{w}$ is
the normalized vertical distance of the claw-hinge from the surface.
Here, $s_{l}/L=1$ and $s_{h}/w=10$. The green dashed line represents
the maximum possible adhesion for the pad. \label{fig:Flex}}
\end{figure}

@@

Based on the three modes of detachment discussed in the previous sections,
one can think of several strategies to detach a fibrillar adhesive pad
from the surface. To illustrate this, let us consider the adhesive
system of a dock beetle. The beetle is known to have 3 sets of hairy
pads in each of their legs, each possessing hairs of different geometries.
To keep our analysis simple, we will assume each leg to have only
two pads, with identical hairs of ``mushroom'' geometry. The distal
and proximal pads possess roughly 500 and 1000 hairs, respectively \cite{RN19}.
Assuming the pads to be rectangular arrays of $20\times25$ and $40\times25$
hairs, we can model this as a one-dimensional system of 20 and 40
hairs, respectively, by combining the hairs along the width. Based
on reported measurements \cite{RN79}, the beetle's ``combined hair''
is thus considered to have an effective pull-off force, $f_{p}=0.5\times25=12.5$
\textgreek{m}N and spring constant, $k_{h}=0.5\times25=12.5$ N m\protect\textsuperscript{-1}.
The beetle's hairs are approximately $l_{h,0}=40$ \textgreek{m}m
long, spaced $w=10$ \textgreek{m}m apart. The adhesive pad also has
a claw, around 200 \textgreek{m}m long, and the leg is connected roughly
at the end of the proximal pad. This will put $\hat{s}_{h}=20$ and
$s_{l}/L=1$, measured relative to the right end of the distal pad.
The beetle's leg is assumed to possess two joints which could serve
as a hinges for rotation (H\textsubscript{1} and H\textsubscript{2}
in Figure \ref{fig:Detachment-pathways} inset). The claw can be used
as an external hinge (H\textsubscript{3}) by flexing the pad around
it.

@@

Inhomogeneous deformation of hairs results in significant reductions
in the adhesion of a fibrillar system. Such a condition can be achieved
by either 1) pulling the pad while maintaining a constant tilt angle,
2) pulling the pad while maintaining a free leg joint or 3) flexing
the pad around the claw. Strategic control of the joint's mobility or claw
can allow the leg to easily switch between the above mechanisms, thus
providing a simple way to reduce adhesion as per necessity. The presence
of a deformable leg can further trigger a spontaneous propagation
of hair detachment due to the leg's elastic recoil, making it suitable
for fast detachment. Arrays with low $\hat{f}_{p}$ and large number
of hairs, with a hair geometry that allows for large deformations
while avoiding buckling and lateral bundling represent the optimal
design conditions to maximize the range of control over adhesion.
The proposed theory is supported by previously reported experimental
observations of leg detachment in dock beetles and highlights possible role of the leg joint and claws to 
enable reversible adhesion. Similar strategies could potentially be adopted in the design of bio-inspired 
artificial fibrillar adhesives to easily switch the adhesion state without the need of asymmetric structures.

@@

We are grateful to Prof. Dr. Hans-J\"{u}rgen Butt and Dr. Thomas Endlein (Max Planck Institute for Polymer Research, Germany) for reviewing the text, and thank Dr. Ren\textcyr{\`\cyre} Hensel
(Leibniz Institute for New Materials, Germany) and
Dr. Bat-El Pinchasik (Tel-Aviv University, Israel) for fruitful discussions. This work was supported by the Deutsche Forschungsgemeinschaft
[Grant number: PI 1351/2-1] and the Max Planck Graduate Center with the Johannes Gutenberg-Universit\"{a}t Mainz [MPGC].
